```{r, setup, include=F}
#rmarkdown::find_pandoc(version = "2.7.3")
source("global.R")
knitr::opts_chunk$set(cache = T, warning=F, message=F, comment="")
global_buffer_force <- F

# library(knitr)
# hook_output = knit_hooks$get('source')  #this is the output for code
# 
# knit_hooks$set(source = function(x, options) {
#   if (!is.null(n <- options$linewidth) & knitr::is_latex_output()) {
#     x <- strwrap(x, width = n, exdent = 4)
#   }
#   hook_output(x, options)
# })
```

\listoffigures

\newpage
# List of Abbreviations
```{r, echo=F}
reactable(
  data.frame(
    "Abbreviation" = c(
      "xts", 
      "IDE",
      "MVP",
      "ITP", 
      "TEV",
      "MSTE",
      "QP"
      
      ),
    "Definition"= c(
      "R-object used to store time series data", 
      "Integrated development environment",
      "Mean variance portfolio",
      "Index tracking problem", 
      "Tracking error variance",
      "Mean square tracking error",
      "Quadratic programming"
      )
  ), 
  style = list(fontFamily="arial", fontSize=13)
) %>% 
  html_save()
```


# Abstract {-}
Asset allocation aims to construct applicable portfolios that follow a particular investment strategy, such as index tracking or optimal risk return. These investment strategies are formulated as equivalent constrained optimization problems whose solutions represent an optimal portfolio depending on the strategy. In practice, these problems are often difficult to solve without limitations because the solution must be discrete or the constraints are too complex for common solvers. If one excludes limitations, the need for metaheuristics such as particle swarm optimization (PSO) arises because they have the ability to find optimal solutions to arbitrarily complex problems. In this thesis, common asset allocation problems are first explained and continuously solved to generate benchmark problems. After introducing the PSO, it is tested using the benchmark problems. Subsequently, the PSO and its variants are used to solve optimization problems that are not analytically solvable. Finally, the most promising variant of the PSO is used to solve a discrete index tracking problem with consideration of transaction costs and a rebalancing constraint. The backtests indicate that the PSO can yield time-stable results for such problems.

# Introduction
In the past, only active portfolio management existed, where the portfolio manager actively bought and sold assets in an attempt to create a portfolio with the client's objective in mind. This approach requires a high level of expertise of the portfolio manager, who must be informed about the financial world and each asset in his portfolio. The goal of this approach is to take advantage of assets that are traded on the market at the wrong price in order to make a profit. However, it is not trivial to spot these false prices, and it is becoming increasingly difficult as the market seems to become more efficient. An efficient market is defined by the fact that the quoted prices of assets reflect all the information about them and thus the exploitation of false prices is no longer possible. This theory has led to passive portfolio management, which is a quantitative approach that focuses more on the portfolio as a whole and less on each individual asset. This approach has proven successful, which is why there has been an increasing shift towards passive portfolio management in recent years. In 2014, 18% of all funds were passively managed, and this share has already risen to 25% in 2018 [@TaDe2019]. The reason for this is complex: On the one hand, passive management is characterized by the fact that many of the quantitative strategies are purely data-driven, which makes it easier to automate a large part of the tasks, thus saving costs as fewer portfolio managers can manage more portfolios. In addition, passively managed portfolios are often characterized by less rebalancing than actively managed portfolios, which keeps transaction costs lower. In addition to the lower cost of passive portfolios, they are also often less risky, which is of great interest to many investors.


A portfolio manager who follows a quantitative strategy and wants to build up his portfolio in a cost-efficient and fully automated way faces a big challenge, because the resulting optimization problems often cannot be solved analytically, e.g. because they are discrete problems. In order to solve such problems anyway, metaheuristics like PSO are used. A metaheuristic is a higher-level optimization method, which considers the actual optimization problem as a black box and must only be able to evaluate the objective function. The disadvantages are that no quality statement can be made about the solution and that the runtime strongly depends on the time needed to evaluate the function, since this has to be done very often. For this reason it is important to test a methaheuristic intensively. It has been shown that the PSO is one of the best methaheuristics for solving portfolio optimization problems that cannot be solved analytically \@ref(BuAd2021). For this reason, the application of PSO to financial mathematical problems is the focus of this thesis. Furthermore, the aim is to develop a deep understanding of the behavior of PSO in order to enable the reader to apply PSO himself. This is supported by the fact that the code for this work is freely available and all data can be downloaded for free via yahoo-finance.


Another goal of this thesis is to introduce the reader to the programming language R, as I think it is perfectly suited for such purposes. A big advantage of R is that it is user-friendly and you can formulate even complicated things in a simple way. I think R deserves to be much more in the public spotlight because it has many advantages over other programming languages like Python.


The following chapters begin by describing basic information about the software used and then address the free availability of the data, which together should ensure reproducibility. After that the mathematical basics and conventions are explained in chapter \@ref(mathfundations). This is followed by an analysis of active and passive investing in chapter \@ref(activevspassive), which among other things describes the basic advantages of passive investing. In the subsequent chapter, two common problems of passive investing are described, the index tracking problem and the optimal return risk portfolio problem. These problems serve as test problems for the PSO, for which first the optimal solution must be obtained by means of analytical solvers. This is done in chapter \@ref(analyticalsolver), additionally the limitations of such analytical solvers are addressed. This is followed by the core topic of this thesis, which starts with the description of the standard PSO and analyzes its behavior. The test problems of the previous chapter are used to test the quality of the standard PSO. At the end of this chapter a discrete problem is explained and solved by the standard PSO, which would not be solvable analytically. In the following chapter \@ref(psovariants), different variations of the PSO are tested by solving a discrete problem, as in the previous chapter. The standard PSO is used as a benchmark. The results indicate that the local PSO and the self-adjusting velocity PSO are most suitable for such problems. In the last chapter \@ref(reallife) the self-adjusting velocity PSO variant is used to solve a real life index tracking problem, which takes into account the transaction costs and the rebalancing constraint. The backtest is promising and suggests that the PSO can be used for such complicated problems.







